---
title: "Paper"
author: "Erin Xu"
date: "2025-11-14"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage[margin=1in]{geometry}
  - \linespread{1.5}
fontsize: 12pt
bibliography: references.bib
nocite: '@*'
link-citations: true
urlcolor: blue
---
jhkjhkjhkjhkjh
spacing
spacing
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents
sdfuhsdkfjhsdfhdsfkjhdsfkj
This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:fgdhfkdjhdsfkhfsdhjhfsjfkdh

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References 